---
title: "Assignment 1"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
The following document includes Assignment 1 code and results

The following libraries are included for this assignment

* **readr** : to read dataset from a url
* **dplyr** : for dataset manapulation
* **rmarkdown** : for markdown

```{r include=FALSE}
library(readr)
library(dplyr)
library(rmarkdown)
```

We will be using the Open Covid dataset 
for state of Ohio by The COVID Tracking project. 
See covidtracking.com for more details about the dataset.

```{r include=FALSE}
data <- read_csv("https://covidtracking.com/data/download/ohio-history.csv")
```

```{r include=FALSE}
# Summary of all the columns
#summary(data)
#View(data)
```

Convert the table into dataframe so it is easy to work with.
```{r include=FALSE}
as.data.frame(data)
```


## Print descriptive statistics for quantitative variable 
To print descriptive statistics we will be using the package dplyr: https://dplyr.tidyverse.org/reference/summarise_all.html

For the sake of readability we will be using the pipe notation `%>%` heavily used by dplyr.
We pipe/chain function  to select the columns death and hospitalized and calulate their mean, min and max.

Heavylifting is done by the `summarise_at` function it takes the following arguments:

* Columns to be summarized e.g. "death", "hospitalized"
* Functions to be used to summarize e.g. mean, min and max
* Should NA values be filtered out (TRUE)

```{r echo = FALSE}
data %>% 
  summarise_at(c("death", "hospitalized"), list(min = min, max = max, mean = mean), na.rm = TRUE)
```

## Print descriptive statistics for categorical variable 
Here we only print all the data that have lower quality grade (B)

```{r echo = FALSE}
data %>% 
  filter(dataQualityGrade == 'B')
```

## Transform variable
Get mean for columns death and hospatilized, grouped by their data quality.
Here we use the `group_by` function that groups the reslting summary by the column `dataQualityGrade`

```{r echo = FALSE}
data %>%
  group_by(dataQualityGrade)%>% 
  summarise(Death = mean(death, na.rm=TRUE),
            Hospitalized = mean(hospitalized, na.rm=TRUE))
```

## Plot deaths vs days only for data with quality A+
Create an intermediate dataframe to plot the results. 
This dataframe is created by filtering the original data to only get results for A+ quality data.
Then select only the columns `date` and `death`

```{r include=FALSE}
filtered_Data <- data %>% 
  filter(dataQualityGrade == 'A+') %>%
  select(date, death)
#head(filtered_Data)
```

This function plots the previously created dataframe.
```{r}
plot(filtered_Data)
```

## References
- https://covidtracking.com/
- https://uc-r.github.io/dplyr

