---
title: "Naive Bayes Classification."
output:
  pdf_document:
    toc: yes
  always_allow_html: true
  html_notebook:
    highlight: textmate
    theme: cerulean
  html_document:
    df_print: paged
    toc: yes
---
***
# Assignment 3 (Naive Bayes Classification.)

Objective here is to use Naive Bayes for classification.

***

## Code Prep

Load given data.
```{r include=FALSE}
#install.packages('readr')
library(readr)
# Data loading
data <- read_csv("UniversalBank.csv", )
```

Sample data
```{r}
head(data)
```

Filter data to only include predictors
```{r}
library(dplyr)
filtered_data = select(data, c(`Personal Loan`, Online, CreditCard))
head(filtered_data)
#tail(filtered_data)
```

Convert the predictors to categorical variables
```{r}
filtered_data$`Personal Loan` <- factor(filtered_data$`Personal Loan`)
filtered_data$Online <- factor(filtered_data$Online)
filtered_data$CreditCard <- factor(filtered_data$CreditCard)
```

### Data Partition
Partition the data into training (60%) and validation (40%) sets.
```{r}
library(caret)
library(ggplot2)
set.seed(13)
train_index = createDataPartition(filtered_data$`Personal Loan`, p=0.6, list=FALSE) # 60% training data
# Train Data (60%)
train_data = filtered_data[train_index,]
valid_data = filtered_data[-train_index,] # rest of the data for validation
```


## Solution A
Pivot table for the training data with Online as a column variable, CC as a row variable, and Loan as a secondary row variable. 
Good info: https://seananderson.ca/2013/10/19/reshape/

### Pivot table using cast and melt 
This pivot table is not used for subsequent calculations but more for my own reference. 
```{r}
#install.packages('reshape')
library(reshape2)
# melt the data first 
# collapse the dataframe retaining "Online" and bringing "CreditCard" and "Personal Loan" from columns to rows

train_data.melt <- melt(data = train_data, id.vars = c("CreditCard", "Personal Loan"))
#head(train_data.melt)
train_data.dcast <-  dcast(train_data.melt, CreditCard + `Personal Loan` ~ variable, fun.aggregate = length)
train_data.dcast

train_data.melt2 <- melt(data = train_data, id.vars = c("CreditCard", "Online"))
#head(train_data.melt2)
train_data.dcast <-  dcast(train_data.melt2, CreditCard + Online ~ variable)
train_data.dcast

with(train_data, table(CreditCard + Online, `Personal Loan`))
#table(train_data)
```

### Pivot table using pivottabler
Used for subsequent calculations. A Pivot table with `Online` as a column variable, `CC` as a row variable, and `Loan` as a secondary row variable.
Note: pivottabler docs: https://cran.r-project.org/web/packages/pivottabler/vignettes/v00-vignettes.html
```{r}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(train_data)
pt$addColumnDataGroups("Online")
pt$addRowDataGroups("CreditCard")
pt$addRowDataGroups("Personal Loan", expandExistingTotals=TRUE)
pt$defineCalculation(calculationName="Total", summariseExpression="n()")
pt$renderPivot()
```

The pivot table above represents training data, the values in the table convey the total count of online account for different combinations of CreditCard and Personal Loan


## Solution B
The probability of loan acceptance when CreditCard = 1 and Online = 1 is = 49/3000 = **0.0163**

## Solution C
### Loan (rows) as a function of Online (columns) 

```{r}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(train_data)
pt$addColumnDataGroups("Online")
pt$addRowDataGroups("Personal Loan")
pt$defineCalculation(calculationName="Total", summariseExpression="n()")
pt$renderPivot()
```
### Loan (rows) as a function of CC (Columns)

```{r}
library(pivottabler)
pt <- PivotTable$new()
pt$addData(train_data)
pt$addColumnDataGroups("CreditCard")
pt$addRowDataGroups("Personal Loan")
pt$defineCalculation(calculationName="Total", summariseExpression="n()")
pt$renderPivot()
```

## Solution D

### P(CC = 1 | Loan = 1) (the proportion of credit card holders among the loan acceptors)

= 86/285 = **0.301**

### P(Online = 1 | Loan = 1)

= 165/285 = **0.589**

### P(Loan = 1) (the proportion of loan acceptors)

= 285/3000 = **0.095**

### P(CC=1|Loan=0)

= 807/2715 = **0.297**

### P(Online = 1 | Loan = 0)

= 1637/2715 = **0.602**

### P(Loan = 0)

= 2715/3000 = **0.905**

## Solution E
Use the quantities computed above to compute the naive Bayes probability P(Loan = 1 | CC = 1, Online = 1)

Using Naive Bayes formula:
`P(Loan = 1 | CC = 1, Online = 1) = P(Loan = 1) [P(CC = 1 | Loan = 1) * P(Online = 1 |Loan = 1 )] / P(Loan = 1) [P(CC = 1 | Loan = 1) * P(Online = 1 |Loan = 1 )] + P(Loan = 0) [P(CC = 1 | Loan = 0) * P(Online = 1 |Loan = 0 )]`

`P(Loan = 1 | CC = 1, Online = 1) = 0.095 ( 0.301 * 0.589) / 0.095 ( 0.301 * 0.589) + 0.905 ( 0.297 * 0.602 )`

`P(Loan = 1 | CC = 1, Online = 1) = 0.0168 / 0.0168 + 0.162`
 
`P(Loan = 1 | CC = 1, Online = 1) = 0.0168 / 0.179 = 0.0938`

**P(Loan = 1 | CC = 1, Online = 1) = 0.0938**


## Solution F

Comparing value calculated using Naive Bayes formula = 0.0938 and value calculated using pivot table = 0.0163

The one calculated from the pivot table appears to be the most correct estimate given the assumptions involved with Naive Bayes. But, looking at both the values Naive Bayes appears to be closer to the actual estimate.

## Solution G

Which of the entries in this table are needed for computing `P(Loan = 1 | CC = 1, Online = 1)`?
The entries needed for computing `P(Loan = 1 | CC = 1, Online = 1)` are:

* `P(Loan = 1)` 
* `P(CC = 1 | Loan = 1)`
* `P(Online = 1 |Loan = 1 )`
* `P(Loan = 0)` 
* `P(CC = 1 | Loan = 0)` 
* `P(Online = 1 | Loan = 0 )`

### Naive Bayes
```{r}
library(e1071)
# or the formula could be Personal Loan` ~ Online + CreditCard
train.nb <- naiveBayes(`Personal Loan` ~ ., data = train_data)
train.nb
```
This naive bayes model is similar to the ones we calculated using the Pivot table in Solution D (e.g. P(Loan = 1) = **0.095** and P(Loan = 0) = **0.905**)

```{r}
# Predict the default status of test dataset 
valid.predict.class <- predict(train.nb, valid_data)
valid.predict.prob <- predict(train.nb, valid_data, type = "raw")
library("gmodels")
# Show the confusion matrix of the classifier
CrossTable(x=valid_data$`Personal Loan`, y=valid.predict.class, prop.chisq = FALSE) 
```


### Entry that corresponds to P(Loan = 1 | CC = 1, Online = 1)

```{r}
test.df <- data.frame(CreditCard = c("1"), Online = c("1"))
test.predict.class <- predict(train.nb, test.df)
test.predict.class
test.predict.prob <- predict(train.nb, test.df, type = "raw")
test.predict.prob

```
Looking at the probabilities, we see that value for `P(Loan = 1 | CC = 1, Online = 1)` = **0.09204853**
which is very similar to the probability we got from Naive Bayes formula = **0.0938** (Solution E) and value calculated using pivot table = **0.0163** (Solution B)
